Parametricity for Primitive Nested Types and GADTs

### 2021-01-12 free-theorems-nesteds/20-mscs.tex
final LMCS submission

\title[Parametricity for Primitive Nested Types and
  GADTs]{Parametricity for Primitive Nested Types}

@@

This paper extends the conference paper~\cite{jgj21} of the same name
by including more exposition, proofs of all theorems, and more
examples. In addition, the discussion of parametricity for GADTs in
Section~\ref{sec:GADTs} is entirely new.

@@

\begin{defi}\label{def:RT-functor}
An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
\begin{itemize}
\item $H^1$ and $H^2$ are functors from $[\set^k,\set]$ to $[\set^k,\set]$
\item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
\item For all $\overline{R : \rel(A,B)}$,
  $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
  \delta^1)_{\overline{A}}$ and
  $\pi_2((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^2
  \delta^2)_{\overline{B}}$
\item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
  (H^1F^1,\,H^2F^2,\,H^*(F^1,F^2,F^*))$
\item The action of $H$ on morphisms is given by
  $H\,(\delta^1,\delta^2) = (H^1\delta^1,H^2\delta^2)$ for
  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
\end{itemize}
\end{defi}
Since the results of applying an endofunctor $H$ to $k$-ary relation
transformers and morphisms between them must again be $k$-ary relation
transformers and morphisms between them, respectively,
Definition~\ref{def:RT-functor} implicitly requires that the following
three conditions hold:
\begin{itemize}
\item $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 \ol{B})$
  whenever $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$
\item $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha},
  H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k) \in \Homrel(R_k,S_k)$
\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
  $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in H^*(F^1,F^2,F^*)\ol{R}$
\end{itemize}
  Note, however, that this last condition is
  automatically satisfied because it is implied by the third bullet
  point of Definition~\ref{def:RT-functor}.

@@

\begin{lem}\label{lem:graph-reln-functors}
If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
transformation, then $\graph{\alpha}$ is in $RT_k$.
\end{lem}
\proof
Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
$\graph{\alpha}^* : [\rel^k,\rel]$. Now, let $\overline{R :
  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
  \beta') : R \to S}$. We want to show that there exists a morphism
$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
such that the diagram on the left below commutes. Since
$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
  S}$ such that each diagram in the middle commutes.
Moreover, since both $h_{\overline{C \times D}} \circ
F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
on the right commute, they must be equal.
%\begin{figure*}[ht]
%\vspace*{-0.1in}
%  \hspace*{-0.85in}
  \begin{minipage}[b]{0.25\linewidth}
 {\small    \[
    \begin{tikzcd}
        \alpha^\wedge\overline{R}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
        \ar[d, "{\epsilon}"']
        & F\overline{A} \times G\overline{B}
        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
        \alpha^\wedge\overline{S}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
        & F\overline{C} \times G\overline{D}
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\small    \[
    \begin{tikzcd}
        R_i
        \ar[d, "{\gamma_i}"']
        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
        &A_i \times B_i
        \ar[d, "{\beta_i \times \beta'_i}"] \\
        S_i
        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
        &C_i \times D_i
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\footnotesize \[
      \begin{tikzcd}[row sep = large]
          F\overline{C}
          &F\overline{C} \times F\overline{D}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &F\overline{D}
          \ar[r, "{\alpha_{\ol{D}}}"]
          &G\overline{D}\\
          &F(\overline{A \times B})
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
      \end{tikzcd}
      \]}
\end{minipage}
%\end{figure*}

@@

\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{\ol K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
\ar[ur, bend left, "{\mathit{Lan}_{\ol K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\gamma}"]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]

@@

An alternative presentation of characterizes the left Kan extension
$(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the adjunction between
natural transformations from $F$ to $G \circ \ol{K}$ and natural
transformations from $\mathit{Lan}_{\ol K} F$ to $G$, for which
$\eta$ is the unit. If Agda were to support a primitive
$\mathsf{Lan}$ for left Kan extensions, we could use this adjunction
to rewrite the type of each data constructor for the GADT
$\mathtt{Seq}$ to arrive at the following equivalent representation:
{\small
\[\begin{array}{l}
\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
\hspace*{0.4in}\mathtt{spair\;:\;(\mathsf{Lan}_{\lambda C\,D.\,C \times D}
  (\lambda C\,D.\,Seq\;C\times Seq\;D))\;A \rightarrow Seq\;A}\\ 
\hspace*{0.4in}\mathtt{sseq\;:\;(\mathsf{Lan}_{\lambda C.\,Nat \to C}
  (\lambda C.\,Nat \rightarrow Seq\;C))\;A \rightarrow
  Seq\;A}
\end{array}\]}
\noindent
\!\!Our calculus will represent $\mathtt{Seq}$ and other GADTs in
precisely this way.

A third representation of left Kan extensions in locally presentable
categories is given in terms of colimits. Writing ${\mathcal C}_0$ for
the full subcategory of finitely presentable objects in the locally
presentable category $\mathcal C$, the left Kan extension can be
expressed as
\[
(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{\ol{X : \,{\mathcal
        C}_0},\, \ol{f : K{\ol X} \to A}}{F \ol X}\]
Thus, if $F : \set^k \to \set$, $\ol{K} :
  \set^k \to \set^h$, $\ol{A} : \set^h$, and $\set_0$ is the full
  subcategory of finitely presentable objects in $\set$, --- i.e., is
  the category finite sets --- we  have that
\begin{equation}\label{eq:colim-form}
(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
      K\ol{S} \to  A}}{F \ol{S}}
\end{equation}
If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
  \ol{f}} : F \ol{S} \to (\mathit{Lan}_{\ol K} F) \ol{A}$ be the morphism
indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
in Equation~\ref{eq:colim-form}.
%Then for any $h$-tuple of functions
%$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
%is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
%$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
%$\ol{f : K\ol{S} \to A}$,\,
%\begin{equation}\label{eq:cocone-def}
%(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
%    \ol{g\, \circ f}}
%  \end{equation}
%holds, where
For any $h$-tuple of functions
$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_{\ol K
  }F) \ol{g}$
is the unique function from $(\mathit{Lan}_{\ol K} F) \ol{A}$ to
$(\mathit{Lan}_{\ol K} F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
$\ol{f : K\ol{S} \to A}$,\,
\begin{equation}\label{eq:cocone-def}
(\mathit{Lan}_{\ol K}F) \ol{g} \circ j_{\ol{S},\, \ol{f}}
= j'_{\ol{S},\, \ol{g\, \circ f}}
: F \ol{S} \to (\mathit{Lan}_{\ol K}F) \ol{B}
\end{equation}
holds, where $j$ is the morphism mapping the cocone into
$(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
    K\ol{S} \to A}}{F \ol{S}}$ and $j'$ is the morphism mapping the
cocone into $(\mathit{Lan}_{\ol K} F) \ol{B} = \colim{\ol{S} : \set_0^k,
  \,\ol{f : K\ol{S} \to B}}{F \ol{S}}$.  Moreover, if $\alpha : F \to
F'$ is a natural transformation, then $\mathit{Lan}_{\ol K} \alpha :
(\mathit{Lan}_{\ol K} F) \to (\mathit{Lan}_{\ol K} F')$ is defined to be the
induced natural transformation whose component $(\mathit{Lan}_{\ol K}
\alpha) \ol{A} : (\mathit{Lan}_{\ol K} F) \ol{A} \to (\mathit{Lan}_{\ol K} F')
\ol{A}$ is the unique function such that $(\mathit{Lan}_{\ol K} \alpha)
\ol{A} \,\circ\, j_{\ol{S}, \ol{f}} = j'_{\ol{S}, \ol{f}} \,\circ\,
\alpha_{\ol{S}}$, where $j$ is the cocone into $(\mathit{Lan}_{\ol K} F)
\ol{A}$ and $j'$ is the cocone into $(\mathit{Lan}_{\ol K} F') \ol{A}$. We
can similarly represent left Kan extensions of functors over $\rel$ in
terms of colimits. In that setting we will denote the morphism mapping
the cocone into the colimit $\iota$. We will make good use of both of
these representations in Section~\ref{sec:term-sem} below.

@@

Intuitively, $\Lan^{\ol{\alpha^0}}_{\ol K} F$ is a syntactic representation
of the left Kan extension of the functor in the variables in
$\ol\alpha$ denoted by $F$ along the functor in the variables in
$\ol\alpha$ denoted by $\ol K$. Using $\Lan$-types, we can therefore
represent the GADT $\mathtt{Seq}$ in our calculus as 
 \begin{align*}
\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. 
\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
\gamma) \right) \beta \right) \alpha
\end{align*}
\noindent
where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
\to C$ abbreviates $(\Lan^\alpha_{\mathtt{Nat}} 1)\,C$.  This
representation of $\mathit{Seq}$ corresponds exactly to the rewriting
in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
  K} F$ allowing extensions along vectors of functors as depicted
above makes it possible to represent GADTs with two or more type
arguments that depend on one another. Such GADTs cannot be represented
using just unary $\Lan$-types, i.e, $\Lan$-types of the form
$(\Lan^{\ol{\alpha^0}}_{\ol K} F)\ol A$.

@@

\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
      K}\rho {[\ol{\alpha := A} ]}}"{left}] && \set \\  
&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\lambda \ol A.\,
      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
\end{tikzcd}\]
and, for all $\ol{N : RT_k}$ and $d : \setsem{\Gamma;\emptyset \vdash
  \Delta}$,
\[\begin{array}{l}
\mu_{\ol{N}\, d} : \lambda \ol{A}.\,
\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
\hspace*{0.8in}\to \lambda \ol B.\, \setsem{\Gamma; 
  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
\end{array}\]
is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
\end{equation}
as given by the universal property of the left Kan extension.  The
relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
  G}$ are defined entirely analogously.

>>>>

\title[Parametricity for Primitive Nested Types and
  GADTs]{Parametricity for Primitive Nested Types and GADTs}

@@

This paper extends the conference paper~\cite{jgj21} by including more
exposition, proofs of all theorems, and more examples. In addition,
the discussion of parametricity for GADTs in Section~\ref{sec:GADTs}
is entirely new.

@@

\begin{defi}\label{def:RT-functor}
An endofunctor $H$ on $RT_k$ is a triple $H = (H^1,H^2,H^*)$, where
\begin{itemize}
\item $H^1$ and $H^2$ are functors from $[\set^k,\set]$ to $[\set^k,\set]$
\item $H^*$ is a functor from $RT_k$ to $[\rel^k,\rel]$
\item For all $\overline{R : \rel(A,B)}$,
  $\pi_1((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^1
  \delta^1)_{\overline{A}}$ and
  $\pi_2((H^*(\delta^1,\delta^2))_{\overline{R}}) = (H^2
  \delta^2)_{\overline{B}}$
\item The action of $H$ on objects is given by $H\,(F^1,F^2,F^*) =
  (H^1F^1,\,H^2F^2,\,H^*(F^1,F^2,F^*))$
\item The action of $H$ on morphisms is given by
  $H\,(\delta^1,\delta^2) = (H^1\delta^1,H^2\delta^2)$ for
  $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$
\end{itemize}
\end{defi}
Since the results of applying an endofunctor $H$ to $k$-ary relation
transformers and morphisms between them must again be $k$-ary relation
transformers and morphisms between them, respectively,
Definition~\ref{def:RT-functor} implicitly requires that the following
three conditions hold:
\begin{itemize}
\item $H^*(F^1,F^2,F^*) \ol{R} : \rel(H^1F^1 \ol{A}, H^2F^2 \ol{B})$
  whenever $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$
\item $H^*(F^1,F^2,F^*)\, \ol{(\alpha, \beta)} = (H^1F^1\ol{\alpha},
  H^2F^2 \ol{\beta})$ whenever $(\alpha_1, \beta_1) \in
  \Homrel(R_1,S_1),..., (\alpha_k, \beta_k)$ $\in \Homrel(R_k,S_k)$
\item If $(\delta^1,\delta^2) : (F^1,F^2,F^*)\to (G^1,G^2,G^*)$ and
  $R_1:\rel(A_1,B_1),...,R_k:\rel(A_k,B_k)$, then
  $((H^1\delta^1)_{\ol{A}}x, (H^2\delta^2)_{\ol{B}}y) \in
  H^*(G^1,G^2,G^*)\ol{R}$ whenever $(x, y) \in H^*(F^1,F^2,F^*)\ol{R}$
\end{itemize}
  Note, however, that this last condition is
  automatically satisfied because it is implied by the third bullet
  point of Definition~\ref{def:RT-functor}.

@@

\begin{lem}\label{lem:graph-reln-functors}
If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
transformation, then $\graph{\alpha}$ is in $RT_k$.
\end{lem}
\proof
Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
$\graph{\alpha}^* : [\rel^k,\rel]$. Let $\overline{R :
  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
  \beta') : R \to S}$. We want to show that there exists a morphism
$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
such that the diagram on the left below commutes. Since
$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
  S}$ such that each diagram in the middle commutes.
Moreover, since both $h_{\overline{C \times D}} \circ
F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
on the right commute, they must be equal.
%\begin{figure*}[ht]
%\vspace*{-0.1in}
%  \hspace*{-0.85in}
  \begin{minipage}[b]{0.25\linewidth}
 {\small    \[
    \begin{tikzcd}
        \alpha^\wedge\overline{R}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
        \ar[d, "{\epsilon}"']
        & F\overline{A} \times G\overline{B}
        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
        \alpha^\wedge\overline{S}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
        & F\overline{C} \times G\overline{D}
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\small    \[
    \begin{tikzcd}
        R_i
        \ar[d, "{\gamma_i}"']
        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
        &A_i \times B_i
        \ar[d, "{\beta_i \times \beta'_i}"] \\
        S_i
        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
        &C_i \times D_i
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\footnotesize \[
      \begin{tikzcd}[row sep = large]
          F\overline{C}
          &F\overline{C} \times F\overline{D}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &F\overline{D}
          \ar[r, "{\alpha_{\ol{D}}}"]
          &G\overline{D}\\
          &F(\overline{A \times B})
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
      \end{tikzcd}
      \]}
\end{minipage}
%\end{figure*}

@@

\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{\ol K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{\ol K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
\end{tikzcd}
\]

@@

An alternative presentation characterizes the left Kan extension
$(\mathit{Lan}_{\ol K} F, \eta)$ in terms of the adjunction between
natural transformations from $F$ to $G \circ \ol{K}$ and natural
transformations from $\mathit{Lan}_{\ol K} F$ to $G$, for which
$\eta$ is the unit. If Agda were to support a primitive
$\mathsf{Lan}$ for left Kan extensions, we could use this adjunction
to rewrite the type of each data constructor for the GADT
$\mathtt{Seq}$ to arrive at the following equivalent representation:
{\small
\[\begin{array}{l}
\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow Seq\;A}\\
\hspace*{0.4in}\mathtt{spair\;:\;(\mathsf{Lan}_{\lambda C\,D.\,C \times D}
  (\lambda C\,D.\,Seq\;C\times Seq\;D))\;A \rightarrow Seq\;A}\\ 
\hspace*{0.4in}\mathtt{sseq\;:\;(\mathsf{Lan}_{\lambda C.\,Nat \to C}
  (\lambda C.\,Nat \rightarrow Seq\;C))\;A \rightarrow
  Seq\;A}
\end{array}\]}
\noindent
\!\!Our calculus will represent $\mathtt{Seq}$ and other GADTs in
precisely this way.

A third representation of left Kan extensions in locally presentable
categories is given in terms of colimits. Writing ${\mathcal C}_0$ for
the full subcategory of finitely presentable objects in the locally
presentable category $\mathcal C$, the left Kan extension can be
expressed as
\[
(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{\ol{X : \,{\mathcal
        C}_0},\, \ol{f : K{\ol X} \to A}}{F \ol X}\]
Thus, if $F : \set^k \to \set$, $\ol{K} :
  \set^k \to \set^h$, $\ol{A} : \set^h$, and $\set_0$ is the full
  subcategory of finitely presentable objects in $\set$, --- i.e., is
  the category of finite sets --- we  have that
\begin{equation}\label{eq:colim-form}
(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
      K\ol{S} \to  A}}{F \ol{S}}
\end{equation}
If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
  \ol{f}} : F \ol{S} \to (\mathit{Lan}_{\ol K} F) \ol{A}$ be the morphism
indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
in Equation~\ref{eq:colim-form}.
%Then for any $h$-tuple of functions
%$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
%is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
%$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
%$\ol{f : K\ol{S} \to A}$,\,
%\begin{equation}\label{eq:cocone-def}
%(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
%    \ol{g\, \circ f}}
%  \end{equation}
%holds, where
For any $h$-tuple of functions
$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_{\ol K
  }F) \ol{g}$
is the unique function from $(\mathit{Lan}_{\ol K} F) \ol{A}$ to
$(\mathit{Lan}_{\ol K} F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
$\ol{f : K\ol{S} \to A}$,\,
\begin{equation}\label{eq:cocone-def}
(\mathit{Lan}_{\ol K}F) \ol{g} \circ j_{\ol{S},\, \ol{f}}
= j'_{\ol{S},\, \ol{g\, \circ f}}
: F \ol{S} \to (\mathit{Lan}_{\ol K}F) \ol{B}
\end{equation}
holds, where $j$ is the morphism mapping the cocone into
$(\mathit{Lan}_{\ol K} F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
    K\ol{S} \to A}}{F \ol{S}}$ and $j'$ is the morphism mapping the
cocone into $(\mathit{Lan}_{\ol K} F) \ol{B} = \colim{\ol{S} : \set_0^k,
  \,\ol{f : K\ol{S} \to B}}{F \ol{S}}$.  Moreover, if $\alpha : F \to
F'$ is a natural transformation, then $\mathit{Lan}_{\ol K} \alpha :
(\mathit{Lan}_{\ol K} F) \to (\mathit{Lan}_{\ol K} F')$ is defined to be the
induced natural transformation whose component $(\mathit{Lan}_{\ol K}
\alpha) \ol{A} : (\mathit{Lan}_{\ol K} F) \ol{A} \to (\mathit{Lan}_{\ol K} F')
\ol{A}$ is the unique function such that $(\mathit{Lan}_{\ol K} \alpha)
\ol{A} \,\circ\, j_{\ol{S}, \ol{f}} = j'_{\ol{S}, \ol{f}} \,\circ\,
\alpha_{\ol{S}}$, where $j$ is the cocone into $(\mathit{Lan}_{\ol K} F)
\ol{A}$ and $j'$ is the cocone into $(\mathit{Lan}_{\ol K} F') \ol{A}$. We
can similarly represent left Kan extensions of functors over $\rel$ in
terms of colimits. In that setting we will denote the morphism mapping
the cocone into the colimit $\iota$. We will make good use of both of
these representations in Section~\ref{sec:term-sem} below.

@@

Intuitively, $\Lan^{\ol{\alpha^0}}_{\ol K} F$ is a syntactic representation
of the left Kan extension of the functor in the variables in
$\ol\alpha$ denoted by $F$ along the functor in the variables in
$\ol\alpha$ denoted by $\ol K$. Using $\Lan$-types, we can therefore
represent the GADT $\mathtt{Seq}$ in our calculus as 
 \begin{align*}
\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. 
\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
\gamma) \right) \beta \right) \alpha
\end{align*}
\noindent
where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
\to C$ abbreviates $(\Lan^\alpha_{\mathtt{Nat}} 1)\,C$.  This
representation of $\mathit{Seq}$ corresponds exactly to the rewriting
in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
  K} F$ allowing extensions along vectors of functors as depicted
above makes it possible to represent GADTs with two or more type
arguments that depend on one another. Such GADTs cannot be represented
using just unary $\Lan$-types, i.e, $\Lan$-types of the form
$(\Lan^{\ol{\alpha^0}}_K F)\ol A$.

@@

\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
      K}\rho {[\ol{\alpha := A} ]}}\;\;\;\;"{left}] && \set \\  
&\set^{|\ol K|} \ar[ur, "\;\;\mathit{Lan}_{\ol{\lambda \ol A.\,
      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
\end{tikzcd}\]
and, for all $\ol{N : RT_k}$ and $d : \setsem{\Gamma;\emptyset \vdash
  \Delta}$,
\[\begin{array}{l}
\mu_{\ol{N}\, d} : \lambda \ol{A}.\,
\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
\hspace*{0.8in}\to \lambda \ol B.\, \setsem{\Gamma; 
  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
\end{array}\]
is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
\end{equation}
as given by the universal property of the left Kan extension.  The
relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
  G}$ are defined entirely analogously.
